Testing for measurement invariance with respect to an ordinal variable

### 2012-07-10 ordinal/ordinal.Rnw
Additions to ordinal paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@152 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{Researchers are often interested in testing for 
  measurement invariance with respect to an ordinal auxiliary variable such as
  age group, income class, or school grade.  In a factor-analytic
  context, these tests are traditionally carried out via a likelihood
  ratio statistic that compares a model where parameters differ across groups
  to a model where parameters are equal across groups.  This test
  neglects the fact that the auxiliary variable is ordinal, and it
  is also known to be overly sensitive at large sample sizes.  In this
  paper, we propose a new measurement invariance test statistic that
  explicitly accounts for the ordinality of the auxiliary variable.
  The statistic is derived from a family of tests based on stochastic
  processes that have recently received attention in the psychometric
  literature.  The statistic is illustrated via an application
  involving real data, and its performance is studied via simulation.
}

@@

The test statistic proposed here is derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{MerZei12}.  % TODO Add kopf
In the following section, we provide an overview of the family and
describe the proposed statistic in detail.  Next, we report on the
results of two simulation studies designed to compare the proposed
test statistic to existing statistics for testing measurement
invariance.  Next, we illustrate the proposed statistic using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the test's use in practice.

@@

It is of interest to test a hypothesis of 
measurement invariance with respect to subsets of model parameters.
The hypothesis can be written as
%% TODO equation
which reflects invariance in the full set of model parameters
$\theta$.  It is straightforward to test a subset of model parameters,
and we do so later in the paper.  In this section, we first provide
background on the family of tests from which the proposed test
statistic is obtained.  We then describe the proposed statistic for
testing measurement invariance with respect to an ordinal variable.

The ordinal test statistic proposed here, along with
related ``non-ordinal'' statistics reported elsewhere, rely on the
casewise contributions to the model likelihood function (i.e., the
{\em scores}).  These contributions are defined as
%% TODO equation
the sum of which equals $\bm{0}$ when maximum likelihood estimates
$\hat{\bm{\theta}}$ are substituted for $\bm{\theta}$.  
Under the hypothesis of measurement invariance, the
individual scores are expected to fluctuate around zero.  Conversely,
when measurement invariance is violated, the scores should
systematically stray from zero for some individuals.
%% TODO add citations:
As described in Merkle and Zeileis (\citeyearNP{MerZei12}; see XX for
further detail), we can create formal test statistics based on these
ideas.  These statistics rely on ordering of individuals with
respect to an auxiliary variable, then taking a cumulative sum of the
scores:
%% TODO insert cumsum
Under %eqref
terms in this cumulative sum combine to follow a Brownian bridge.
Thus, we can summarize the empirical cumulative scores in various
manners, obtaining critical values and $p$-values from the Brownian
bridge.

@@

\subsection{An Ordinal Test Statistic}

@@

The statistic that we proposed here is defined as
%% TODO insert equation
so that all individuals with the same value of the ordinal variable
simultaneously enter into the cumulative sum.  The test statistic is
then the maximum of this cumulative sum.  Closed-form solutions for
this particular functional of a Brownian bridge are unavailable, but
critical values and $p$-values can be obtained through repeated
simulation of a Brownian bridge.  This functionality is built in to
R's strucchange package \cite{strucchange}, which we use to carry out
the tests for general SEMs.

@@

In this simulation, we demonstrate that the proposed test statistic is
sensitive to ordinal measurement invariance violations, moreso than
other competing statistics.

@@

@@

>>>>

\abstract{Researchers are often interested in testing for 
  measurement invariance with respect to an ordinal auxiliary variable such as
  age group, income class, or school grade.  In a factor-analytic
  context, these tests are traditionally carried out via a likelihood
  ratio test statistic comparing a model where parameters differ across groups
  to a model where parameters are equal across groups.  This test
  neglects the fact that the auxiliary variable is ordinal, and it
  is also known to be overly sensitive at large sample sizes.  In this
  paper, we propose test statistics that
  explicitly account for the ordinality of the auxiliary variable.
  The statistics are derived from a family of tests based on stochastic
  processes that have recently received attention in the psychometric
  literature.  The statistics are illustrated via an application
  involving real data, and their performance is studied via simulation.
}

@@

The test statistic proposed here is derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{KopZei10,MerZei12}.
In the following section, we provide an overview of the family and
describe the proposed statistics in detail.  Next, we report on the
results of two simulation studies designed to compare the proposed
test statistics to existing tests of measurement
invariance.  Next, we illustrate the proposed statistics using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the tests' use in practice.

@@

We will consider situations in which a $p$-dimensional variable $X$
with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
model with density $f(\bm{x}_i; \bm{\theta})$ and
associated joint log-likelihood
\begin{equation} \label{eq:loglik}
  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
\end{equation}
where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
the distribution.

We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
from a structural equation model with assumed multivariate normality,
though the proposed tests extend beyond this family of models.  Under
the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
$\bm{\theta}$ can 
be estimated by maximum likelihood (ML), i.e.,
\begin{equation} \label{eq:ml}
  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
\end{equation}
or equivalently by solving the first order conditions
\begin{equation}
    \label{eq:ml1}
  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
\end{equation}
where 
\begin{equation}
  \label{eq:score}
  {\bm s}({\bm \theta}; x_i) ~=~ \left(
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
    \dots,
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
  \right)^\top,
\end{equation}
is the score function of the model (the partial derivative of the casewise
likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
\dots, n$ essentially measures the extent to which the model maximizes
each individual's likelihood: as an individual's scores go further
from zero, the model provides a poorer description of that individual.

Measurement invariance tests essentially test the assumption that all
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
hypothesis is typically tested against an alternative
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $v_i$ reflects individual $i$'s value on an auxiliary variable
$V$ and $\nu$ is a threshold dividing individuals into groups based on
$V$.  In this paper, we generally focus on situations where $V$ is
ordinal and 
where $\nu$ is unknown.  In the section below, we review tests where
$V$ is continuous (and $\nu$ is unknown) as background for our
proposed tests for ordinal $V$.

\subsection{Tests for Continuous $V$}
% stopped here

As described in Merkle and Zeileis (\citeyearNP{MerZei12}; see XX for
further detail), we can create formal test statistics based on these
ideas.  These statistics rely on ordering of individuals with
respect to an auxiliary variable, then taking a cumulative sum of the
scores:
%% TODO insert cumsum
Under %eqref
terms in this cumulative sum combine to follow a Brownian bridge.
Thus, we can summarize the empirical cumulative scores in various
manners, obtaining critical values and $p$-values from the Brownian
bridge.

@@

\subsection{Proposed Tests for Ordinal $V$}

@@

In contrast to the likelihood ratio test, the statistics that we
propose here explicitly take into 
account the fact that the auxiliary variable is ordinal.  They are
similar to the statistics from % eqref
above, except that we focus on ``bins'' of individuals at each level of the
ordinal variable:
%% TODO insert equation
Critical values associated with these test statistics can be obtained
by applying these functionals to bins of a Brownian bridge, where the
bin sizes equal the empirical proportion of individuals at each level
of the ordinal variable.
Closed-form solutions for
this particular functional of a Brownian bridge are unavailable, but
critical values and $p$-values can be obtained through repeated
simulation of a Brownian bridge.  This functionality is built in to
R's strucchange package \cite{strucchange}, which can be used to carry out
the tests for general SEMs.

@@

In this simulation, we demonstrate that the proposed test statistics are
sensitive to ordinal measurement invariance violations, moreso than
other competing statistics.  We generate data from a two-factor,
six-indicator model, with a measurement invariance violation occurring
in the three unique variances associated with the first factor.  We
use the proposed test statistics to test for measurement invariance
simultaneously across all six unique variances, mimicking the first
step of the ``prescribed'' course of measurement invariance tests. % cite

For comparison, we also compute two statistics that treat the ordinal
auxiliary variable as categorical.  These include the likelihood ratio
test of measurement 
invariance in the six unique variance parameters, along with .

\subsection{Method}
As described above, data were generated from a two-factor model
lacking measurement invariance in three of six unique variance
parameters.  Magnitude of measurement invariance violation, sample
size, and number of categories of the ordinal variable were manipulated.
We examined three sample sizes ($n=120, 480, 960$), three numbers of
categories ($l=4, 8, 12$), and seven magnitudes of invariance
violations.  These violations started at level $1+l/2$ and continued
for higher levels of the ordinal variable.  The unique variances for
the ``violating'' levels deviated from the lower levels' unique
variances by $d$~times the parameters' asymptotic standard
errors (scaled by $\sqrt{n}$), with $d=0, 0.25, 0.5, \dots, 1.5$.

For each combination of $n \times l \times d$, 5,000 datasets were
generated and tested via the XX statistics described above.  In all
conditions, we maintained equal sample sizes at each level of the
ordinal variable.

%% sweave code here

\subsection{Results}

@@

\subsection{Method}

\subsection{Results}

@@

%% should we care about non-ordinal violations when using an ordinal variable?
